Neural Power Units

### 2020-10-16 paper/main.tex

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage[preprint]{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{todonotes}

\usepackage{bm}
\usepackage{amsmath}
\usepackage{graphicx}

\graphicspath{{./../../plots/}}
\newcommand{\real}{\text{real}}

\title{Bayesian Neural Arithmetic Units -- Leveraging The Complex Plane}
\author{
  Niklas Heim,
  V\'aclav \v Sm\'idl,
  Tom\'a\v s Pevn\'y \\
  Artificial Intelligence Center\\
  Czech Technical University\\
  Prague, CZ 120 00\\
  \texttt{\{niklas.heim, vasek.smidl, tomas.pevny\}@aic.fel.cvut.cz}\\
}

@@

\begin{abstract}
  Common Neural Networks poorly approximate simple arithmetic operations.  The
  new Neural Arithmetic Units aim to overcome this difficulty, but are limited
  either limited to operate on positive numbers (NALU by
  \citet{trask_neural_2018}), or can only represent simple addition and
  multiplication (NAU \& NMU by \citet{madsen_neural_2020}).
  We present the first Neural Arithmetic Unit that operates on the full domain
  of real numbers and is capable of learning arbitrary power functions.
  For a theoretically grounded sparsification of the units we employ Bayesian
  compression on the model weights via Automatic Relevance Determination.
\end{abstract}

@@

It is known that Neural Networks can approximate functions arbitrarily
well\todo{cite universal approx}.  However, this strength of approximation
often comes at the cost of generalization beyond the training data manifold.
\todo{mention ODEs as use case}

@@

\citet{trask_neural_2018} have demonstrated how severe this problem is even for
the simplest arithmetic operations, such as summing or multiplying two numbers.
In order to increase the power of abstraction of Neural Networks they propose a
\emph{Neural Arithmetic Logic Unit} (NALU) which can help to overcome this
problem \cite{trask_neural_2018} and is capable of learning the arithmetic
addition, subtraction, multiplication, and division
$\{+,-,\times,\div\}$ with stunning extrapolation accuracy.  However,
the NALU comes with the severe limitation not being able to handle negative
inputs due to the logarithm in the multiplication part of the NALU:

\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm W& = \tanh(\hat{\bm W}) \odot \sigma(\hat{\bm M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm W(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}

A multiplication layer that can handle negative inputs was introduced by \citet{madsen_neural_2020}.
The \emph{Neural Multiplication Unit} (NMU) is defined by Eq.~\ref{eq:nmu} and is typically used in
conjunction with the so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i M_{ij} z_{i} + 1 - M_{ij}  &M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm A \bm x &A_{ij}=\min(\max(A_{ij}, 0), 1)
\end{align}
In both NMU and NAU the weights are clipped to $[0,1]$, and typically regularized
with $\mathcal{R}$:
\begin{align}
  \label{eq:rsparse}
  \mathcal{R} = \sum_{ij} \min(W_{ij}, 1-W_{ij})
\end{align}

The combination of NAU and NMU can thus learn $\{+,-,\times\}$, but no division.

\todo{put this in acknowledgements
all the results in this paper were create with the help of the following Julia
packages: \cite{rackauckas_differentialequationsjl_2017} + (Flux.jl)}

\section{Neural Power Unit}%
\label{sec:neural_power_unit}

Our \emph{Neural Power Unit} (NPU) can learn arbitrary power functions (which
includes division) while still being able to correctly deal with negative
inputs.  The NPU layer is defined by
\begin{align}
  k_i &= \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases} \\
  \bm r &= |\bm x| \\
  \bm m &= \exp(\bm M \log(\bm r)) \odot \cos(\bm M \bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the elementwise absolute value, and $\bm M$
the multiplication matrix that we want to learn.

The NPU is inspired by the multiplication part of the NALU
Eq.~\ref{eq:nalu_mult}, extended to negative inputs by taking advantage of the
complex logarithm. In general, for any complex number $z$\footnote{Note that
complex numbers can be represented in the polar, complex plane where
$z=re^{i\varphi}$ with $r>0$ and $\varphi \in (0,2\pi)$.}
\begin{align}
  \log(z) = \log\left(r\cdot e^{i\varphi}\right)
     = \log(r) + i\varphi.
\end{align}
However, as $z$ is assumed to be real, we can simplify as follows:
\begin{align}
  \log(z) = \log(r) + ik\pi,
\end{align}
where $k \in [0,1]$ is zero for positive inputs and one for negative inputs $z$.

\begin{align}
  \bm m &= \exp(\bm M \log(\bm x)) \\
    &= \exp(\bm M ( \log(\bm r) + i\pi\bm k ))
\end{align}

\begin{align}
  \bm m_{\text{re}} &= \real(\exp(\bm M \log \bm r + i\pi\bm M\bm k)) \\
    &= \real(\exp(\bm M \log \bm r) \odot (\cos(\pi \bm M \bm k) + i \sin(\pi \bm M \bm k))) \\
    &= \exp(\bm M \log \bm r) \odot \cos(\pi \bm M \bm k)
\end{align}

\section{Experimetns}%
\label{sec:experimetns}

\subsection{10 param func}%
\label{sub:10_param_func}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-task-nmu-paper.pdf}
%   \caption{MSE optimization result}%
%   \label{fig:10-param-func-task-nmu-paper}
% \end{figure}
% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-bayes-task-nmu-paper.pdf}
%   \caption{ARD result with MSE starting point}%
%   \label{fig:10-param-func-bayes-task-nmu-paper}
% \end{figure}
% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-bayes-task-nmu-paper-history.pdf}
%   \caption{ARD history}%
%   \label{fig:history}
% \end{figure}
% 
% \subsection{10 param func with sqrt and power}%
% \label{sub:10_param_func_with_sqrt_and_power}
% 
% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-task-sqrt-power.pdf}
%   \caption{10-param-func-task-sqrt-Power}%
%   \label{fig:10-param-func-task-sqrt-power}
% \end{figure}
% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-bayes-task-sqrt-power.pdf}
%   \caption{10-param-func-bayes-task-sqrt-Power}%
%   \label{fig:10-param-func-bayes-task-sqrt-power}
% \end{figure}
% \begin{figure}
%   \centering
%   \includegraphics[width=0.8\linewidth]{10-param-func-bayes-task-sqrt-power-history.pdf}
%   \caption{10-param-func-bayes-task-sqrt-Power-history}%
%   \label{fig:10-param-func-bayes-task-sqrt-power-history}
% \end{figure}

\subsection{Gradient surfaces}%
\label{sub:gradient_surfaces}

>>>>

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage[final]{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools}

\newtheorem*{remark}{Remark}

\newtheorem*{definition}{Definition}

\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

% tikz
\usepackage{tikz}
\usetikzlibrary{%
  arrows,%
  backgrounds,
  intersections,
  shapes,
  shapes.misc,% wg. rounded rectangle
  shapes.arrows,%
  chains,%
  matrix,%
  positioning,% wg. " of "
  scopes,%
  decorations.pathmorphing,% /pgf/decoration/random steps | erste Graphik
  shadows,%
  patterns% for hatched rect
}

% plots
\graphicspath{{./../plots/}}
\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{polar}
\usepgfplotslibrary{smithchart}
\usepgfplotslibrary{statistics}
\usepgfplotslibrary{dateplot}

\newcommand{\real}{\text{Re}}
\newcommand{\imag}{\text{Im}}
\newcommand{\wre}{w_{r}}
\newcommand{\wim}{w_{i}}
\newcommand{\Wre}{\bm W_{r}}
\newcommand{\Wim}{\bm W_{i}}
\newcommand{\npu}{\text{NPU}}
\newcommand{\naivenpu}{\text{NaiveNPU}}
\newcommand{\nau}{\text{NAU}}
\newcommand{\nmu}{\text{NMU}}
\newcommand{\nalu}{\text{NALU}}
\newcommand{\inalu}{\text{iNALU}}
\newcommand{\dense}{\text{Dense}}
\newcommand{\chain}{\text{Chain}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% tables
\usepackage{booktabs}
\renewcommand{\arraystretch}{1.2}

\title{Neural Power Units}
\author{
  Niklas Heim,
  Tom\'a\v s Pevn\'y,
  V\'aclav \v Sm\'idl\\
  Artificial Intelligence Center\\
  Czech Technical University\\
  Prague, CZ 120 00\\
  \texttt{\{niklas.heim, tomas.pevny, vasek.smidl\}@aic.fel.cvut.cz}\\
}

@@

\begin{abstract}
  Conventional Neural Networks can approximate simple arithmetic operations,
  but fail to generalize beyond the range of numbers that were seen during
  training.  \emph{Neural Arithmetic Units} aim to overcome this difficulty,
  but current arithmetic units are either limited to operate on positive
  numbers or can only represent a subset of arithmetic operations. We introduce
  the \emph{Neural Power Unit} (NPU)\footnote{
    Implementation of Neural Arithmetic Units: \url{github.com/nmheim/NeuralArithmetic.jl}.
    The code to reproduce our experiments is available at \url{github.com/nmheim/NeuralPowerUnits}.
  }
  that operates on the \emph{full domain of
  real numbers} $\mathbb{R}$ and is capable of \emph{learning arbitrary power functions} in
  a single layer.  The NPU thus fixes the shortcomings of existing arithmetic
  units and extends their expressivity. We achieve this by using
  complex arithmetic without requiring a conversion of the network to complex
  numbers $\mathbb{C}$.
  A simplification of the unit to the \emph{RealNPU} yields a highly transparent model.
  We show that the NPUs outperform their competitors in terms of
  accuracy and sparsity on artificial arithmetic datasets, and that the RealNPU can discover
  the governing equations of a dynamical system only from data.
\end{abstract}

@@

Numbers and simple algebra are essential not only to human intelligence but
also to the survival of many other
species~\citep{dehaene_number_2011,gallistel_finding_2018}. A successful,
intelligent agent should, therefore, be able to perform simple arithmetic.  State
of the art neural networks are capable of learning arithmetic, but they
fail to extrapolate beyond the ranges seen during
training~\citep{suzgun_evaluating_2018,lake_generalization_2018}.  The
inability to generalize to unseen inputs is a fundamental problem that hints at
a lack of \emph{understanding} of the given task. The model merely memorizes
the seen inputs and fails to abstract the true learning task.  The failure of
numerical extrapolation on simple arithmetic tasks has been shown
by~\cite{trask_neural_2018}, who also introduced a new class of \emph{Neural
Arithmetic Units} with good extrapolation performance on some arithmetic
tasks.

Including Neural Arithmetic Units in standard neural networks promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially important for tasks in which
the data generating process contains mathematical relationships.  They also
promise to reduce the number of parameters needed for a given
task, which can improve the explainability of the model.
We demonstrate this in a \emph{Neural Ordinary Differential Equation} (NODE,
\cite{chen_neural_2019}), where a handful of neural arithmetic units can
outperform a much bigger network built from dense layers (Sec.~\ref{sub:fractional_sir_model}).
Moreover, our new unit can be used to directly read out the correct generating ODE
from the fitted model.  This is in line with recent efforts to build
\emph{transparent} models instead of attempting to explain black-box models
\citep{rudin_stop_2019}, like conventional neural networks.
We refer to the terminology by \cite{lipton_mythos_2017} which defines the
potential of understanding the parameters of a given model as \emph{transparency by decomposability}.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. It significantly increases its complexity, and we
observe only a slight improvement in performance.
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range of real
numbers, converge much more reliably, but cannot represent
division.

\subsection*{Our Contributions}%
\label{sub:our_contribution}

\textbf{Neural Power Unit.}
We introduce a new arithmetic layer (NPU, Sec.~\ref{sec:neural_power_unit})
which is capable of learning products of power functions $(\prod x_{i}^{w_i})$ of
arbitrary real inputs $x_i$ and power $w_i$, thus including multiplication
($x_1\times x_2 = x_1^1 x_2^1$) as well as division $(x_1\div
x_2=x_1^{1}x_2^{-1})$. This is achieved by using formulas from complex
arithmetic (Sec.~\ref{sub:naive_npu}).  Stacks of NAUs and NPUs can thus learn
the full spectrum of simple arithmetic operations.

\textbf{Convergence improvement.}
We address the known convergence issues of neural arithmetic units by
introducing a \emph{relevance gate} that smooths out the loss surface of the
NPU (Sec.~\ref{sub:npu}). With the relevance gate, which helps to learn to
ignore variables, the NPU reaches extrapolation errors and sparsities that are on
par with the NMU on ($\times$) and outperforms NALU on ($\div,\sqrt{\cdot}$).

\textbf{Transparency.}
We show how a power unit can be used as a highly transparent\footnote{as defined by
\cite{lipton_mythos_2017}} model for equation discovery
of dynamical systems. Specifically, we demonstrate its ability to identify
a model that can be interpreted as a SIR model with fractional powers
(Sec.~\ref{sub:fractional_sir_model}) that was used to fit the COVID-19
outbreak in various countries \citep{taghvaei_fractional_2020}.

@@

Several different approaches to automatically solve arithmetic tasks have
been studied in recent years. Approaches include Neural GPUs
\citep{kaiser_neural_2016}, Grid LSTMs \citep{kalchbrenner_grid_2016}, Neural
Turing Machines \citep{graves_neural_2014}, and Neural Random Access Machines
\citep{kurach_neural_2016}. They solve tasks like binary addition and
multiplication, or single-digit arithmetic. The Neural Status Register
\citep{faber_neural_2020} focuses on control flow.  The Neural Arithmetic
Expression Calculator \citep{chen_neural_2018}, a hierarchical reinforcement
learner, is the only method that solves the division problem, but it operates on
character sequences of arithmetic expressions. Related is symbolic integration
with transformers \citep{lample_deep_2019}. Unfortunately, most of the named models
have severe problems with extrapolation \citep{madsen_measuring_2019,saxton_analysing_2019}.
A solution to the extrapolation problem could be Neural Arithmetic Units. They are
designed with an inductive bias towards systematic, arithmetic computation.
However, currently, they are limited in their capability of expressing the full
range of simple arithmetic operations ($+,\times,\div$).
In the following two sections, we briefly describe the currently available
arithmetic layers, including their advantages and drawbacks.

\subsection{Neural Arithmetic Logic Units}%
\label{sub:neural_arithmetic_logic_unit}

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  To increase the power of
abstraction for arithmetic tasks, they propose the \emph{Neural Arithmetic
Logic Unit} (NALU), which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:

\begin{definition}[{\bf NALU}]
  The NALU consists of a ($+$) and a ($\times$) path with shared weights $\bm W$ and $\bm M$.
  \begin{align}
    \label{eq:nalu_add}
    \text{Addition: }       & \bm a = \bm{\hat W} \bm x
                            & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}) \\
    \label{eq:nalu_mult}
    \text{Multiplication: } & \bm m = \exp (\bm{\hat W}\log(|\bm x|+\epsilon)) & &\\
    \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                            & \bm g& = \sigma(\bm G\bm x)
  \end{align}
  with inputs $\bm x$ and learnt parameters $\bm W$, $\bm M$, and $\bm G$.
\end{definition}

Additionally, the logarithm destabilizes training to the extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$) and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU, and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through several mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetic.

\subsection{Neural Multiplication Unit \& Neural Addition Unit}%
\label{sub:neural_multiplication_unit}

Instead of trying to fix the NALU's convergence issues,
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.

\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
  \begin{align}
    \label{eq:nmu}
    \text{NMU: } &y_j = \prod_i \hat M_{ij} x_{i} + 1 - \hat M_{ij}  & \hat M_{ij}=\min(\max(M_{ij}, 0), 1)\\
    \label{eq:nau}
    \text{NAU: } &\bm y = \bm{\hat{A}} \bm x & \hat A_{ij}=\min(\max(A_{ij}, -1), 1)
  \end{align}
  with inputs $\bm x$, and learnt parameters $\bm M$ and $\bm A$.
\end{definition}

Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$, and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

\section{Neural Power Units}%
\label{sec:neural_power_unit}

To fix the deficiencies of current arithmetic units, we propose a new
arithmetic unit (inspired by NALU) that can learn arbitrary products of power functions $(\prod x_{i}^{w_i})$ (including
$\times,\div$) for positive and negative numbers, and still train well.
Combined with the NAU, we solve the full range of arithmetic operations.
This is possible through a simple modification of the ($\times$)-path of the
NALU (Eq.~\ref{eq:nalu_mult_2}). 
We suggest to replace the logarithm of the absolute value by the complex logarithm and
to allow $\bm W$ to be complex as well. Since the complex logarithm is defined for
negative inputs, the NPU does not have a problem with negative numbers. A complex $\bm W$ 
improves convergence at the expense of transparency (see
Sec.~\ref{sub:fractional_sir_model}). The improvement during training might be
explained by the additional imaginary parameters that make it possible to avoid
regions with an uninformative gradient signal.

\subsection{Naive Neural Power Unit -- NaiveNPU}%
\label{sub:naive_npu}

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp (\bm W\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm y = \exp(\bm W\log \bm x) = \exp\left((\Wre + i\Wim) \log\bm x\right),
\end{align}
where the input $\bm x$ is still a vector of real numbers.
With the polar form for a complex number $z=re^{i\theta}$ the complex log applied
to a real number $x=re^{ik\pi}$ is
\begin{align}
  \log x = \log r + ik\pi,
\end{align}
where $k=0$ if $x\geq0$ and $k=1$ if $x<0$.
Using the complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$,
resulting in a layer that can process both positive and negative numbers correctly.
A complex weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
outputs, we can avoid this doubling by considering only the real part
of the output $\bm y$:
\begin{align}
  \real(\bm y) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula $e^{ix} = \cos x + i\sin x$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.

\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit, with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers, is defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm y = \exp(\Wre \log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x| + \epsilon,
    \quad
    k_i = \begin{cases}
       0 & x_i \geq 0 \\
       1 & x_i < 0
    \end{cases},
  \end{gather}
  with inputs $\bm x$, machine epsilon $\epsilon$, and learnt parameters $\Wre$ and $\Wim$.
\end{definition}

\begin{figure}
  \begin{minipage}{.45\textwidth}
      \resizebox{0.9\textwidth}{!}{\input{../tikz/npu-diagram.tex}}
      \captionof{figure}{NaiveNPU diagram, with input $\bm x$ and output $\bm y$.
      Vectors in green, trainables in orange, functions in blue.}
      \label{fig:npu_diagram}
  \end{minipage}
  \hspace{0.01\textwidth}
  \begin{minipage}{0.54\textwidth}
    \resizebox{\textwidth}{!}{\input{../tikz/gatednpu-diagram.tex}}
    \captionof{figure}{NPU diagram. The NPU has a relevance gate $\bm g$
    (hatched background) in front of the input to the unit to prevent
    zero gradients.}
    \label{fig:gatednpu_diagram}
  \end{minipage}
\end{figure}

\subsection{The Relevance Gate -- NPU}%
\label{sub:npu}

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the function
$f:\mathbb{R}^2 \rightarrow \mathbb{R}$,
which is the identity on one of two inputs. The task is defined by the loss $\mathcal L$:
\begin{gather}
  \mathcal{L} = \sum_i |m(x_1,x_2) - f(x_1,x_2)| = \sum_i |m(x_1,x_2) - x_{1,i}|,\nonumber\\
  \label{eq:grad_surf}
  \text{where } m = \naivenpu \text{ with } (\Wre,\Wim) \in \mathbb{R}^{1\times 2} \\
  \text{and } x_1 \sim \mathcal U(0,2), x_2 \sim \mathcal U(0,0.05). \nonumber
\end{gather}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm $\mathcal{G}$
\begin{equation}
  \mathcal{G}(\Wre) = \norm{\frac{\partial\mathcal{L}}{\partial\Wre}}_2
\end{equation}
of the NaiveNPU for a batch of two-dimensional inputs.
Even in this simple example, the gradient of the NaiveNPU is close to
zero in large parts of the parameter space. This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of our layer that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.

\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm y = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot (|\bm x|+\epsilon) + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \geq 0 \\
      \hat g_i & x_i < 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1),
  \end{gather}
  with inputs $\bm x$, and learnt parameters $\Wre$, $\Wim$ and $\bm g$.
\end{definition}

\begin{figure}
  \centering
  \includegraphics[width=.6\textwidth]{npu_gatednpu_id_grad.pdf}
  \caption{Gradient norm $\mathcal G$ of NaiveNPU and NPU for the task of learning
  the identity on $x_1$ (black areas are beyond the color scale). Inputs and
  loss are defined in Eq.~\ref{eq:grad_surf}. The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  in the central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient norm $\mathcal G$ of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as $L_1$. In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as a transparent model.

\textbf{Initialization}%
\label{sub:initialization}
We recommend initializing the NPU with a Glorot Uniform distribution on the
real weights $\Wre$. The imaginary weights $\Wim$ can be initialized to zeros,
so they will only be used where necessary, and the gate $\bm g$ with 0.5, so
the NPU can choose to output 1.

\begin{definition}[\textbf{RealNPU}]
  In many practical tasks, such as multiplication or division, the final value
  of $\bm W_i$ should be equal to zero. We will denote NPU with removed
  parameters for the imaginary part as RealNPU and study the impact of this
  change on convergence in Sec.~\ref{sec:experiments}.
\end{definition}

\section{Experiments}%
\label{sec:experiments}

In Sec.~\ref{sub:fractional_sir_model}, we show how the NPU can help to build
better NODE models. Additionally, we use the RealNPU as a highly transparent
model, from which we can directly recover the generating equation of an ODE
containing fractional powers.  Subsequent
Secs.~\ref{sub:simple_arithmetic_task} and \ref{sub:large_scale_arithmetic_task}
compare the NPU to prior art (NALU and NMU) on arithmetic tasks typically used
to benchmark arithmetic units.

\subsection{A Step Towards Equation Discovery of an Epidemiological Model}%
\label{sub:fractional_sir_model}

Data-driven models such as \emph{SINDy}~\citep{champion_data-driven_2019} or
\emph{Neural Ordinary Differential Equations} (NODE, \cite{chen_neural_2019})
are used more and more in scientific applications.
Recently, \emph{Universal Differential Equations} (UDEs,
\cite{rackauckas_universal_2020}) were introduced which aim to combine
data-driven models with physically informed differential equations to maximize
interpretability/explainability of the resulting models.

If an ODE model is composed of dense layers, its direct interpretation is
problematic and has to be performed retrospectively. The class of models based on
SINDy is transparent by design, however it can only provide explanation
within a linear combination of predefined set of basis functions. Thus, it
cannot learn models with unknown fractional powers. 
With this experiment we aim to show that the NPU can \emph{potentially} discover
exact ODE models\footnote{The demonstration given here is not intended to be
used in practice. For real-world predictions in such a sensitive area much more
post-processing is needed to ensure safe predictions.}.
An example of an ODE that contains powers is a modification of the well-known
epidemiological SIR model \citep{kermack_contribution_1927} to fractional powers
(fSIR, \cite{taghvaei_fractional_2020}), which was shown to be a beneficial
modification for modelling the COVID-19 outbreak.  The SIR model is
built from three variables: $S$ (susceptible), $I$ (infectious), and $R$
(recovered/removed).  Arguably the most important part of the model is the
transmission rate $r$, which is typically taken to be proportional to the
product of $S$ and $I$.  \citet{taghvaei_fractional_2020} argue that,
especially in the initial phase of an epidemic, the boundary areas of infected
and susceptible cells scale with a fractional power, which leads to
Eq.~\ref{eq:fracsir_r}:
\begin{align}
  \label{eq:fracsir}
  \frac{dS}{dt} &= -r(t) + \eta R(t), &
  \frac{dI}{dt} &=  r(t) - \alpha I(t), &
  \frac{dR}{dt} &= \alpha I(t) - \eta R(t),\\
  \label{eq:fracsir_r}
  & & r(t) &= \beta I(t)^\gamma S(t)^\kappa,
\end{align}

We have numerically simulated one realization of the fSIR model with the
parameters $\alpha=0.05$, $\beta=0.06$, $\eta=0.01$, $\gamma=\kappa=0.5$, in 40
time steps that are equally spaced in the time interval $T=(0,200)$, such that
the training data $\bm X = [S_t,I_t,R_t]_{t=1}^{40}$
contains one time series each for $S$, $I$, and $R$.  The initial conditions
$\bm u_0=[S_0,I_0,R_0]$ are set to $S_0=100$, $I_0=0.01$, and $R_0=0$, as shown in Figure
\ref{fig:pareto_sir} (right).
We fit the data with three
different NODEs composed of different model types: a dense network, the NPU, and
the RealNPU. An exemplary model is: $\npu=\chain(\npu(3,h),\nau(h,3))$ with
variable hidden size $h$.  The detailed models are defined in
Tab.~\ref{tab:fsir_models}. The training objective is the loss $\mathcal{L}$
with $L_1$ regularization.
\begin{align}
  \label{eq:sir_loss}
  \mathcal{L} = \text{MSE}(\bm X,\text{NODE}_\theta(\bm u_0)) + \beta||\bm\theta||_1.
\end{align}
We train each model for 3000 steps with the ADAM optimizer and a learning rate
of 0.005, and subsequently with LBFGS until convergence (or for maximum 1000
steps).  For each model type, we run a small grid search to build a Pareto front
with $h\in \{6,9,12,15,20\}$ and $\beta\in\{0,0.01,0.1,1\}$, where each
hyper-parameter pair is run five times. The resulting Pareto front is shown on
the left of Fig.~\ref{fig:pareto_sir}. The NPU reaches much sparser and
better solutions than the dense network.
\begin{figure}
  \centering
  \resizebox{.9\textwidth}{!}{\input{../plots/pareto_sir.tex}}
  \caption{Pareto fronts of the dense network, NPU, and RealNPU. The
  NPU reaches solutions with lower MSE and fewer parameters than the dense
  net. The RealNPU mostly yields worse results than the NPU. Just in a few
  cases it converges to very sparse models with good MSE.}%
  \label{fig:pareto_sir}
\end{figure}
The RealNPU has problems to converge in the majority of cases, however, 
there are a few models in the bottom left that
reach a very low MSE and have very few parameters. The best of these models is
shown in Fig.~\ref{fig:sir_gatednpu_modelps}.  It looks strikingly similar to
the fSIR model in matrix form:
\renewcommand*{\arraystretch}{1.3}
\begin{equation}
  \begin{bmatrix}
    \dot S \\ \dot I \\ \dot R
  \end{bmatrix}
  =
  \begin{bmatrix}
    -\beta & 0 & \eta \\
    \beta & -\alpha & 0 \\
    0 & \alpha & \eta
  \end{bmatrix}
